Towards Comparable Active Learning

### 2024-05-31 doc/al_benchmark_paper.tex
paper

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Why we need 50 restarts}\label{sec:restarts}
To evaluate how many restarts are necessary to obtain conclusive and reproducible results in an AL experiment, we set the number of restarts of the margin sampling baseline for the Splice dataset to 100.
This allows us firstly, to obtain a very strong estimation of the "true" average performance of margin sampling on this dataset and secondly, to draw subsets from this pool of 100 runs.
Setting the size of our draws to $\alpha$ and sampling uniformly, we can approximate a cross-validation process with $\alpha$ restarts.
Each of these draws can be interpreted as a reported result in AL literature where the authors employed $\alpha$ restarts.
%We provide evidence in Fig. \ref{fig:restarts} that previous works might not have evaluated their experiments with a sufficient number of restarts.
Figure \ref{fig:restarts} shows the "true" mean performance of margin sampling (green) in relation to random sampling (black) and the oracle performance (red).
We display 5 random draws of size $\alpha$ in blue.
We can observe that even for a relatively high number of restarts the variance between the samples is extremely high, resulting in some performance curves being worse that random and some being significantly better.
When setting $\alpha = 50$ we observe all samples to converge close to the true mean performance. 
For that reason, we employ 50 restarts of every experiment in this work.
\begin{figure}
        \hspace{-20mm}
	\includegraphics[width=1.24\linewidth]{img/ablation_restarts}
	\caption{Random draws from a pool of 100 runs for margin sampling on the Splice dataset with different numbers of repetitions ($\alpha=\{5,10,50\}$). Green curves signify the true mean performance of all 100 runs, while the samples are blue. Even with 5 or 10 repetitions, we can observe that single draws for margin sampling display below-random performance (black), while the true mean should be above random.}
	\label{fig:restarts}
\end{figure}

@@

To avoid algorithms memorizing these datasets they are generated during the experiment, depending on $s_\D$. \\ [1mm]

@@

	\caption{Our two synthetic datasets.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Classification Model}\label{sec:choosing_the_classifier}
\begin{wraptable}{R}{0.55\textwidth}
    \vspace{-0.7cm}
    \caption{Overview of available batch sizes for each dataset}
    \vspace{0.1cm}
    \label{tab:batch_sizes}
    \begin{tabular}{l|c|c|c|c|c|c|c}
                    & 1 & 5 & 20 & 50 & 100 & 500 & 1000 \\
         \hline
        Enc. DNA    & o & o &&&&& \\
        Enc. Splice & o & o & o & o &&&\\
        TopV2       & o & o & o & o &&& \\
        Splice      & o & o & o & o & o && \\
        DNA         & o & o & o & o & o && \\
        USPS        & o & o & o & o & o && \\
        Enc. USPS   & o & o & o & o & o && \\
        Enc. Cifar10& o & o & o & o & o && \\
        Enc. FMnist & o & o & o & o & o && \\
        News        &&& o & o & o & o &\\
        FMnist      &&&&&& o & o\\
        Cifar10     &&&&&& o & o \\
    \end{tabular}
    \vspace{-0.35cm}
\end{wraptable}
The model $\hat y$ can be trained in two ways. Either the parameters of the model are reset to a fixed initial setting $\hat y^{(0)}$ after each AL iteration and the classifier is trained from scratch with the updated labeled set $\mathcal{L}^{(i)}$, or the previous state $\hat y^{(i-1)}$ is retained and the classifier is fine-tuned on $\mathcal{L}^{(i)}$ for a reduced number of epochs.
In this work, we use the fine-tuning method for un-encoded datasets to save computational time, while we use the from-scratch training for embedded datasets since they have very small classifiers and this approach generally produces better results.
Our fine-tuning scheme always trains for at least one epoch and employs an aggressive early stopping with a patience of 2 afterwards.\\

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Realism vs. Variance}\label{sec:realism}
We would like to point out that some design choices for this framework prohibit direct transfer of our results to practical applications. 
This is a conscious choice, as we think that this is a necessary trade-off between realism and experiment variance.
We would like to highlight the following design decisions: \\
(i) Creating test and validation splits from the full dataset rather than only the labeled seed set. Fully fledged test and validation splits are unobtainable in practice, but they provide not only a better approximation of algorithm performance, but also a better foundation for hyperparameter tuning, which is bound to reduce variance in the experiment. \\
(ii) Choosing smaller classifiers instead of SOTA models. Since we are not interested in archiving a new SOTA in any classification problem, we instead opt to use smaller classifiers for the following reasons:
Smaller classifiers generally exhibit more stable training behavior, on average require fewer sampled datapoints to reach their full-dataset-performance and have faster training times.
For every dataset, the chosen architecture's hyperparameters are optimized to archive maximum full-dataset performance.
Generally, we use MLPs for tabular, RestNet18 for image and BiLSTMs for text datasets.
Every encoded dataset is classified by a single linear layer with softmax activation.
For an overview of architectures and hyperparameters please refer to Appendix \ref{app:hyperparameters}. \\

@@

	\caption{Performances for acquisition functions on each dataset and aggregated for un-encoded and encoded datasets. Performance is shown as average ranks over restarts (1.0 is the best rank). Algorithms are sorted by aggregated performance on un-encoded datasets. For brevity, we only display per-dataset standard deviations for Splice. All standard deviations lie within $[0.001, 0.024]$.}

@@

        %\resizebox{\columnwidth}{!}{

@@

Oracle & 1.0$\pm$0.011 & 1.0 & 1.0 & 3.9 & 7.3 & 1.0 & 1.0 & 2.3 & 1.9 \\
Margin & 3.9$\pm$0.019 & 4.2 & \textbf{2.1} & 4.7 & \textbf{2.9} & \textbf{2.4} & 2.7 & 3.3 & 3.9 \\
Badge & \textbf{3.0$\pm$0.014} & 6.2 & 2.9 & \textbf{3.5} & 3.2 & 3.3 & \textbf{2.5} & 3.5 & 4.9 \\
DSA & 7.1$\pm$0.016 & 7.1 & 7.1 & 3.8 & 3.5 & 5.7 & 5.9 & 5.8 & 6.2 \\
CoreGCN & 6.6$\pm$0.012 & 4.8 & 9.4 & 6.1 & 4.8 & 4.0 & 5.4 & 5.9 & 7.7 \\
BALD & 3.9$\pm$0.011 & 4.6 & 5.3 & 10.6 & 5.8 & 7.2 & 3.8 & 5.9 & 6.9 \\
Entropy & 6.4$\pm$0.018 & \textbf{3.8} & 7.2 & 6.0 & 3.3 & 9.1 & 8.2 & 6.3 & 6.0 \\
LSA & 5.8$\pm$0.009 & 6.6 & 5.2 & 6.1 & 9.2 & 7.0 & 6.0 & 6.6 & 6.9 \\
Coreset & 6.8$\pm$0.01 & 8.8 & 9.6 & 5.2 & 5.3 & 8.0 & 9.4 & 7.6 & 6.6 \\
Random & 8.6$\pm$0.01 & 9.0 & 5.2 & 6.9 & 9.7 & 7.4 & 6.6 & 7.6 & 6.4 \\
TypiClust & 8.3$\pm$0.009 & 9.8 & 11.0 & 6.4 & 9.6 & 11.0 & 10.6 & 9.5 & 8.6

@@

        %}

>>>>

	\resizebox{\columnwidth}{!}{

@@

}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Why we need 50 restarts}\label{sec:restarts}
To evaluate how many restarts are necessary to obtain conclusive and reproducible results in an AL experiment, we set the number of restarts of the margin sampling baseline for the Splice dataset to 100.
This allows us firstly, to obtain a very strong estimation of the "true" average performance of margin sampling on this dataset and secondly, to draw subsets from this pool of 100 runs.
Setting the size of our draws to $\alpha$ and sampling uniformly, we can approximate a cross-validation process with $\alpha$ restarts.
Each of these draws can be interpreted as a reported result in AL literature where the authors employed $\alpha$ restarts.
%We provide evidence in Fig. \ref{fig:restarts} that previous works might not have evaluated their experiments with a sufficient number of restarts.
Figure \ref{fig:restarts} shows the "true" mean performance of margin sampling (green) in relation to random sampling (black) and the oracle performance (red).
We display 5 random draws of size $\alpha$ in blue.
We can observe that even for a relatively high number of restarts the variance between the samples is extremely high, resulting in some performance curves being worse that random and some being significantly better.
When setting $\alpha = 50$ we observe all samples to converge close to the true mean performance. 
For that reason, we employ 50 restarts of every experiment in this work.
\begin{figure}
        \hspace{-20mm}
	\includegraphics[width=1.24\linewidth]{img/ablation_restarts}
	\caption{Random draws from a pool of 100 runs for margin sampling on the Splice dataset with different numbers of repetitions ($\alpha=\{5,10,50\}$). Green curves are the mean performance of all 100 runs, while the samples are blue. Even with 5 or 10 repetitions, we can observe that single draws for margin sampling display below-random performance (black), while the true mean should be above random.}
	\label{fig:restarts}
\end{figure}

@@

To avoid algorithms memorizing these datasets they are generated during the experiment, depending on $s_\D$. \\

@@

	\caption{Synthetic "Honeypot" and "Diverging Sine" datasets. The decision boundary is not part of the dataset and serves only as a visual guide.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Classification Model}\label{sec:choosing_the_classifier}
\begin{wraptable}{R}{0.65\textwidth}
    \vspace{-0.7cm}
    \caption{Overview of available batch sizes for each dataset}
    \vspace{0.1cm}
    \label{tab:batch_sizes}
    \begin{tabular}{l|c|c|c|c|c|c|c|c}
                    & B & 1 & 5 & 20 & 50 & 100 & 500 & 1000 \\
         \hline
        Enc. DNA    & 40 & o & o &&&&& \\
        Honeypot    & 60 & o & o &&&&& \\
        Diverging Sine& 60 & o & o &&&&& \\
        Enc. Splice & 100 & o & o & o & o &&&\\
        TopV2       & 200 & o & o & o & o &&& \\
        Splice      & 400 & o & o & o & o & o && \\
        DNA         & 300 & o & o & o & o & o && \\
        USPS        & 400 & o & o & o & o & o && \\
        Enc. Cifar10& 450 & o & o & o & o & o && \\
        Enc. FMnist & 500 & o & o & o & o & o && \\
        Enc. USPS   & 600 & o & o & o & o & o && \\
        News        & 3K &&& o & o & o & o &\\
        FMnist      & 10K &&&&&& o & o\\
        Cifar10     & 10K &&&&&& o & o \\
    \end{tabular}
    \vspace{-0.35cm}
\end{wraptable}
The model $\hat y$ can be trained in two ways. Either the parameters of the model are reset to a fixed initial setting $\hat y^{(0)}$ after each AL iteration and the classifier is trained from scratch with the updated labeled set $\mathcal{L}^{(i)}$, or the previous state $\hat y^{(i-1)}$ is retained and the classifier is fine-tuned on $\mathcal{L}^{(i)}$ for a reduced number of epochs.
In this work, we use the fine-tuning method for un-encoded datasets to save computational time, while we use the from-scratch training for embedded datasets since they have very small classifiers and this approach generally produces better results.
Our fine-tuning scheme always trains for at least one epoch and employs an aggressive early stopping with a patience of 2 afterwards.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Realism vs. Variance}\label{sec:realism}
We would like to point out that some design choices for this framework prohibit direct transfer of our results to practical applications. 
This is a conscious choice, as we think that this is a necessary trade-off between realism and experiment variance.
We would like to highlight the following design decisions: \\
(i) Creating test and validation splits from the full dataset rather than only the labeled seed set. Fully fledged test and validation splits are unobtainable in practice, but they provide not only a better approximation of algorithm performance, but also a better foundation for hyperparameter tuning, which is bound to reduce variance in the experiment. \\
(ii) Choosing smaller classifiers instead of SOTA models. Since we are not interested in archiving a new SOTA in any classification problem, we instead opt to use smaller classifiers for the following reasons:
Smaller classifiers generally exhibit more stable training behavior, on average require fewer sampled datapoints to reach their full-dataset-performance and have faster training times.
For every dataset, the chosen architecture's hyperparameters are optimized to archive maximum full-dataset performance.
Generally, we use MLPs for tabular, RestNet18 for image and BiLSTMs for text datasets.
Every encoded dataset is classified by a single linear layer with softmax activation.
For an overview of architectures and hyperparameters please refer to Appendix \ref{app:hyperparameters}.

@@

	\caption{Performances for acquisition functions on real-world datasets, aggregated for un-encoded and encoded datasets. Performance is shown as average ranks over restarts (1.0 is the best rank). Algorithms are sorted by aggregated performance on un-encoded datasets. For brevity, we only display per-dataset standard deviations for Splice. All standard deviations lie within $[0.0007, 0.024]$.}

@@

        \resizebox{\columnwidth}{!}{

@@

Oracle 	& 1.0$\pm$0.011 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.0 & 1.9 \\
Margin 	& 3.9$\pm$0.019 & 4.2 & \textbf{2.1} & 4.4 & \textbf{2.9} & \textbf{2.4} & 2.7 & 3.2 & 3.9 \\
Badge  	& \textbf{3.0$\pm$0.014} & 6.2 & 2.9 & \textbf{3.4} & 3.2 & 3.3 & \textbf{2.5} & 3.5 & 4.9 \\
DSA 	& 7.1$\pm$0.016 & 7.1 & 7.1 & 3.6 & 3.5 & 5.7 & 5.9 & 5.7 & 6.2 \\
BALD 	& 3.9$\pm$0.011 & 4.6 & 5.3 & 10.1 & 5.8 & 7.2 & 3.8 & 5.8 & 6.9 \\
CoreGCN & 6.6$\pm$0.012 & 4.8 & 9.4 & 5.7 & 4.8 & 4.0 & 5.4 & 5.8 & 7.7 \\
Entropy & 6.4$\pm$0.018 & \textbf{3.8} & 7.2 & 5.7 & 3.4 & 9.1 & 8.2 & 6.3 & 6.0 \\
LSA 	& 5.8$\pm$0.009 & 6.6 & 5.2 & 5.8 & 8.7 & 7.0 & 6.0 & 6.4 & 6.9 \\
Random 	& 8.6$\pm$0.01 & 9.0 & 5.2 & 6.5 & 9.2 & 7.4 & 6.6 & 7.5 & 6.4 \\
Coreset & 6.8$\pm$0.01 & 8.8 & 9.6 & 4.9 & 5.3 & 8.0 & 9.4 & 7.5 & 6.6 \\
TypiClust & 8.3$\pm$0.009 & 9.8 & 11.0 & 6.0 & 9.1 & 11.0 & 10.6 & 9.4 & 8.6

@@

       	}
